---
title: "Transactions for January 2019"
output:
     pdf_document:
         latex_engine: xelatex
---

```{r echo=FALSE, error=TRUE, message=FALSE}
library(tidyverse)
library(knitr)
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, 
                      message = FALSE, cache.lazy = FALSE)
currentYear <- 2019
currentMonth <- 1
df <- read.csv(file = 'transactions.csv')
df <- df[df$year==currentYear & df$month==currentMonth,]
dfxR <- df[!df$category == 'rent',]

orderByDate <- df[order(df$day),]
orderByPrice <- df[order(df$price),]

totalByCat <- summarise(group_by(df, category), price=sum(price))
totalByCatxR <- summarise(group_by(dfxR, category), price=sum(price))
meanByCat <- summarise(group_by(df, category), price=mean(price))

```

```{r echo=FALSE, message=FALSE}
# Average number of days between each purchase in a certain category. 
daysApart <- function(cat) {
  dG <- df[df$category==cat,'day']
  len <- length(dG)
  diffs <- rep(NA, len-1)
  
  for (i in 1:len) {
    for (j in 1:len) {
      if (i-1 == j) {
        diffs[j] = dG[i] - dG[j]
      }
    }
  }
  
  return(diffs)
}
```

```{r echo=FALSE}
totalSpent <- format(sum(totalByCat$price), nsmall=2)
fixedCosts <- format(sum(c(700,42.94,4.99)), nsmall=2)
variableCosts <- format(sum(totalByCat$price) - sum(c(700,4.99,42.94)), nsmall=2)

timesEatingOut <- as.integer(summary(droplevels(df)$category)['eating out'])
avgCostMeal <- as.numeric(totalByCat[totalByCat$category == 'groceries','price'])/(60-timesEatingOut)
avgCostEatingOut <- as.numeric(meanByCat[meanByCat$category == 'eating out','price'])

daysApartGroceries <- format(mean(daysApart('groceries')), nsmall=2)
OCeatingOut <- avgCostEatingOut / avgCostMeal
```

## Total spent this month: $`r totalSpent` 

### Fixed/recurring costs: $`r fixedCosts`

### Variable costs: $`r variableCosts`
&nbsp;

#### Useful statistics for spendings, excluding rent. 

```{r echo=FALSE}
spendxRstats <- t(data.frame(c(format(summary(dfxR$price), nsmall=2))))
rownames(spendxRstats) <- NULL
kable(spendxRstats)
```
&nbsp;

#### Number of transactions in each category. 

```{r echo=FALSE}
transNums <- t(data.frame(c(summary(droplevels(df)$category))))
rownames(transNums) <- NULL
kable(transNums)
```
&nbsp;

#### Average and total spendings by category. 

```{r echo=FALSE}
avgCosts <- meanByCat[order(-meanByCat$price),]
avgCosts$price <- format(avgCosts$price, nsmall=2)

totalCosts <- totalByCat[order(-totalByCat$price),]
totalCosts$price <- round(totalCosts$price, digits=2)

allCosts <- inner_join(avgCosts, totalCosts, by='category')
colnames(allCosts) <- c('category', 'average', 'total')

kable(allCosts)
```
&nbsp;

#### The average number of days between for grocery runs is `r daysApartGroceries`.
&nbsp;

#### The opportunity cost of eating out is `r OCeatingOut` meals at home. 

\newpage
\center

## Amount of money spent each day, colourized by category, excluding rent. 

```{r echo=FALSE, fig.height=6}
ggplot(dfxR) +
  aes(x=day, y=price, fill=category) +
  geom_bar(stat='identity')
```

\newpage 

## Total spendings by category as a pie chart. 

```{r echo=FALSE, fig.height=4}
ggplot(totalByCat) +
  aes(x=2, y=price, fill=category) +
  geom_bar(stat='identity') +
  coord_polar(theta='y') +
  theme(axis.ticks.y = element_blank(), 
        axis.text.y = element_blank()) +
  labs(x='', y='')
```


## Total spendings by category as a bar chart. 

```{r echo=FALSE, fig.height=4}
ggplot(totalByCat) +
  aes(x=category, y=price, fill=category) +
  geom_bar(stat='identity') +
  theme(axis.text.x = element_text(angle=90, vjust=0.5))
```

\newpage

## Total spendings by category, excluding rent, as a pie chart. 

```{r echo=FALSE, fig.height=4}
ggplot(totalByCatxR) +
  aes(x=2, y=price, fill=category) +
  geom_bar(stat='identity') +
  coord_polar(theta='y') +
  theme(axis.ticks.y = element_blank(), 
        axis.text.y = element_blank()) +
  labs(x='', y='')
```

## Total spendings by category, excluding rent, as a bar chart. 

```{r echo=FALSE, fig.height=4}
ggplot(totalByCatxR) +
  aes(x=category, y=price, fill=category) +
  geom_bar(stat='identity') +
  theme(axis.text.x = element_text(angle=90, vjust=0.5))
```

\newpage

## Scatterplot of all transactions. 

```{r echo=FALSE, fig.height=6}
ggplot(dfxR) +
  aes(x=day, y=price, colour=category) +
  geom_point()
```

